\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-20 paper.qmd
Update the screen shots and videos of simulated data example

Let $h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}$ be the function that maps a point in \kD{} space to it's bin centroid ($C^{(k)}$). This is the model points in \kD{} space.

When $k = 2$, one of the representations of a bin is the hexagonal centroid (@fig-NLDR-scurve (c)).

@@

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. Define the set $H^k$ to be the set of points that belong to bin $k$. We use the \pD{} Euclidean mean of the points in $H^k$ to map the centroid $(h_x^k, h_y^k)$ to a point in \pD{}. Let the $i$-th component of the \pD{} mean be

@@

with corresponding \pD{} vector $\vec{f_i}$. Then, $f(h\circ g)(x)$ maps a \pD{} point to the \pD{} model estimate, completing the model cycle.

@@

<!--langevitour with pD model for S-curve-->
::: {#fig-scurve_sc layout-ncol="3" fig-pos="H"}
![](figures/scurve/sc_1.png){#fig-scurve_sc1}

![](figures/scurve/sc_2.png){#fig-scurve_sc2}

![](figures/scurve/sc_3.png){#fig-scurve_sc3}

@@

To investigate which is the reasonable representation to visualize the five spherical Gaussian cluster data or all NLDR methods provide equivalent information, we visualize all the models in \pD{} space. Models from all NLDR methods show five well-separated clusters (see @fig-gau1_sc, @fig-gau2_sc, @fig-gau3_sc, @fig-gau4_sc, and @fig-gau5_sc). This suggests that for the five Gaussian cluster dataset, all NLDR methods effectively preserve the global structure. tSNE, UMAP, and PHATE display clusters with varying densities, indicating their ability to capture within-cluster variation (see @fig-gau1_sc, @fig-gau2_sc, and @fig-gau3_sc). On the other hand, both TriMAP and PaCMAP show clusters with flat surfaces, suggesting a failure to capture within-cluster variation (see @fig-gau4_sc and @fig-gau5_sc). Therefore, TriMAP and PaCMAP do not capture the local structure as effectively as other methods.

::: {#fig-gau1_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_tsne_1.png){#fig-gau1_sc1}

![](figures/five_gau_clusters/sc_tsne_2.png){#fig-gau1_sc2}

![](figures/five_gau_clusters/sc_tsne_3.png){#fig-gau1_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with tSNE in high-D, a video of the tour animation is available at (<https://youtu.be/oQxEb4wRdHI>).
:::

::: {#fig-gau2_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_umap_1.png){#fig-gau2_sc1}

![](figures/five_gau_clusters/sc_umap_2.png){#fig-gau2_sc2}

![](figures/five_gau_clusters/sc_umap_3.png){#fig-gau2_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with UMAP in high-D, a video of the tour animation is available at (<https://youtu.be/JW49csPpDx4>).
:::

::: {#fig-gau3_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_phate_1.png){#fig-gau3_sc1}

![](figures/five_gau_clusters/sc_phate_2.png){#fig-gau3_sc2}

![](figures/five_gau_clusters/sc_phate_3.png){#fig-gau3_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with PHATE in high-D, a video of the tour animation is available at (<https://youtu.be/xqhRODt8Z-o>).
:::

::: {#fig-gau4_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_trimap_1.png){#fig-gau4_sc1}

![](figures/five_gau_clusters/sc_trimap_2.png){#fig-gau4_sc2}

![](figures/five_gau_clusters/sc_trimap_3.png){#fig-gau4_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with TriMAP in high-D, a video of the tour animation is available at (<https://youtu.be/X-uHSB4VoZ0>).
:::

::: {#fig-gau5_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_pacmap_1.png){#fig-gau5_sc1}

![](figures/five_gau_clusters/sc_pacmap_2.png){#fig-gau5_sc2}

![](figures/five_gau_clusters/sc_pacmap_3.png){#fig-gau5_sc3}

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with PaCMAP in high-D, a video of the tour animation is available at (<https://youtu.be/UxvU8HQd7JM>).
:::

@@

::: {#fig-pbmc1_sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_1.png){#fig-pbmc1_sc1}

![](figures/pbmc3k/sc_2.png){#fig-pbmc1_sc2}

![](figures/pbmc3k/sc_3.png){#fig-pbmc1_sc3}

@@

::: {#fig-pbmc2_sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_4.png){#fig-pbmc2_sc4}

![](figures/pbmc3k/sc_5.png){#fig-pbmc2_sc5}

![](figures/pbmc3k/sc_6.png){#fig-pbmc2_sc6}

@@

::: {#fig-mnist1_sc layout-ncol="3" fig-pos="H"}
![](figures/mnist/sc_1.png){#fig-mnist1_sc1}

![](figures/mnist/sc_2.png){#fig-mnist1_sc2}

![](figures/mnist/sc_3.png){#fig-mnist1_sc3}

>>>>

Let $g : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}$ be the function that maps a point in \kD{} space to it's bin centroid ($C^{(k)}$). This is the model points in \kD{} space.

When $k = 2$, one of the representations of a bin is the hexagonal centroid (@fig-NLDR-scurve (c)). Also, it could be the mean of the data points within each hexagon.

@@

When neighboring centroids are connected with edges in Delaunay triangulation in \gD{}, there are situations where distant centroids are also connected, which generates long edges. To generate a smooth surface in \gD{}, these long edges are removed (@fig-NLDR-scurve (d)).

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. Define the set $H^k$ to be the set of points that belong to bin $k$. We use the \pD{} Euclidean mean of the points in $H^k$ to map the centroid $C_{h}^{(k)} = (c_{h1}, c_{h2})$ to a point in \pD{}. Let the $i$-th component of the \pD{} mean be

@@

with corresponding \pD{} vector $\vec{f_i}$. Then, $f(g(x))$ maps a \pD{} point to the \pD{} model estimate, completing the model cycle.

@@

<!--langevitour with pD model for S-curve-->
::: {#fig-scurve_sc layout-ncol="3" fig-pos="H"}
![](figures/scurve/sc_1.png)

![](figures/scurve/sc_2.png)

![](figures/scurve/sc_3.png)

@@

To investigate which is the reasonable representation to visualize the five spherical Gaussian cluster data or all NLDR methods provide equivalent information, we visualize all the models in \pD{} space. Models from all NLDR methods show five well-separated clusters (see @fig-gau1_sc, @fig-gau2_sc, @fig-gau3_sc, @fig-gau4_sc, and @fig-gau5_sc). This suggests that for the five Gaussian cluster dataset, all NLDR methods effectively preserve the global structure. tSNE displays clusters with varying densities, indicating their ability to capture within-cluster variation (see @fig-gau1_sc). On the other hand, both UMAP, PHATE, PaCMAP and TriMAP show clusters with flat surfaces, suggesting a failure to capture within-cluster variation (see @fig-gau2_sc, @fig-gau3_sc, @fig-gau4_sc and @fig-gau5_sc). Therefore, UMAP, PHATE, PaCMAP and TriMAP do not capture the local structure as effectively as other methods.

::: {#fig-gau1_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_tsne_1.png)

![](figures/five_gau_clusters/sc_tsne_2.png)

![](figures/five_gau_clusters/sc_tsne_3.png)

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with tSNE in \pD{}, a video of the tour animation is available at (<https://youtu.be/RASEE7N5MbM>).
:::

::: {#fig-gau2_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_umap_1.png)

![](figures/five_gau_clusters/sc_umap_2.png)

![](figures/five_gau_clusters/sc_umap_3.png)

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with UMAP in \pD{}, a video of the tour animation is available at (<https://youtu.be/iG4bCPkJilw>).
:::

::: {#fig-gau3_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_phate_1.png)

![](figures/five_gau_clusters/sc_phate_2.png)

![](figures/five_gau_clusters/sc_phate_3.png)

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with PHATE in \pD{}, a video of the tour animation is available at (<https://youtu.be/L_PVLGwfOS0>).
:::

::: {#fig-gau4_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_pacmap_1.png)

![](figures/five_gau_clusters/sc_pacmap_2.png)

![](figures/five_gau_clusters/sc_pacmap_3.png)

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with PaCMAP in \pD{}, a video of the tour animation is available at (<https://youtu.be/z07cKXi8EJQ>).
:::

::: {#fig-gau5_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_trimap_1.png)

![](figures/five_gau_clusters/sc_trimap_2.png)

![](figures/five_gau_clusters/sc_trimap_3.png)

Screen shots of the **langevitour** of the five Gaussian clusters dataset, shows the model with TriMAP in \pD{}, a video of the tour animation is available at (<https://youtu.be/Chs1lYAoX2w>).
:::

@@

::: {#fig-pbmc1_sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_1.png)

![](figures/pbmc3k/sc_2.png)

![](figures/pbmc3k/sc_3.png)

@@

::: {#fig-pbmc2_sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_4.png)

![](figures/pbmc3k/sc_5.png)

![](figures/pbmc3k/sc_6.png)

@@

::: {#fig-mnist1_sc layout-ncol="3" fig-pos="H"}
![](figures/mnist/sc_1.png)

![](figures/mnist/sc_2.png)

![](figures/mnist/sc_3.png)

### 2024-05-20 paper.tex
Update the screen shots and videos of simulated data example

Let \(h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}\) be
the function that maps a point in \kD{} space to it's bin centroid
(\(C^{(k)}\)). This is the model points in \kD{} space.

When \(k = 2\), one of the representations of a bin is the hexagonal
centroid (Figure~\ref{fig-NLDR-scurve} (c)).

@@

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{}
vectors that represent bin centroids. Define the set \(H^k\) to be the
set of points that belong to bin \(k\). We use the \pD{} Euclidean mean
of the points in \(H^k\) to map the centroid \((h_x^k, h_y^k)\) to a
point in \pD{}. Let the \(i\)-th component of the \pD{} mean be

@@

with corresponding \pD{} vector \(\vec{f_i}\). Then, \(f(h\circ g)(x)\)
maps a \pD{} point to the \pD{} model estimate, completing the model
cycle.

@@

\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/scurve/sc_1.png}

\subcaption{\label{fig-scurve_sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/scurve/sc_2.png}

\subcaption{\label{fig-scurve_sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/scurve/sc_3.png}

\subcaption{\label{fig-scurve_sc3}}

\end{minipage}%

@@

@@

\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_1.png}

\subcaption{\label{fig-pbmc1_sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_2.png}

\subcaption{\label{fig-pbmc1_sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_3.png}

\subcaption{\label{fig-pbmc1_sc3}}

\end{minipage}%

@@

\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_4.png}

\subcaption{\label{fig-pbmc2_sc4}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_5.png}

\subcaption{\label{fig-pbmc2_sc5}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/pbmc3k/sc_6.png}

\subcaption{\label{fig-pbmc2_sc6}}

\end{minipage}%

@@

According to Figure~\ref{fig-mnist1_sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP (see
Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \pD{} data. Additionally, the model shows
a twisted pattern within the non-linear structure in \pD{} space (see
Figure~\ref{fig-mnist1_sc2}), which is an additional pattern not visible
in the \gD{} representation (see Figure~\ref{fig-pacmap-author}).
Furthermore, as shown in Figure~\ref{fig-mnist1_sc3}, some long edges
exist in the \pD{} space that are not recognized as long edges in the
\gD{} representation. However, PaCMAP is a \emph{good} \gD{}
representation of MNIST digit 1 data.

@@

\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/mnist/sc_1.png}

\subcaption{\label{fig-mnist1_sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/mnist/sc_2.png}

\subcaption{\label{fig-mnist1_sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\captionsetup{labelsep=none}\includegraphics{figures/mnist/sc_3.png}

\subcaption{\label{fig-mnist1_sc3}}

\end{minipage}%

>>>>

Let \(g : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}\) be
the function that maps a point in \kD{} space to it's bin centroid
(\(C^{(k)}\)). This is the model points in \kD{} space.

When \(k = 2\), one of the representations of a bin is the hexagonal
centroid (Figure~\ref{fig-NLDR-scurve} (c)). Also, it could be the mean
of the data points within each hexagon.

@@

When neighboring centroids are connected with edges in Delaunay
triangulation in \gD{}, there are situations where distant centroids are
also connected, which generates long edges. To generate a smooth surface
in \gD{}, these long edges are removed (Figure~\ref{fig-NLDR-scurve}
(d)).

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{}
vectors that represent bin centroids. Define the set \(H^k\) to be the
set of points that belong to bin \(k\). We use the \pD{} Euclidean mean
of the points in \(H^k\) to map the centroid
\(C_{h}^{(k)} = (c_{h1}, c_{h2})\) to a point in \pD{}. Let the \(i\)-th
component of the \pD{} mean be

@@

with corresponding \pD{} vector \(\vec{f_i}\). Then, \(f(g(x))\) maps a
\pD{} point to the \pD{} model estimate, completing the model cycle.

@@

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/scurve/sc_1.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/scurve/sc_2.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/scurve/sc_3.png}\end{minipage}%

@@

To investigate which is the reasonable representation to visualize the
five spherical Gaussian cluster data or all NLDR methods provide
equivalent information, we visualize all the models in \pD{} space.
Models from all NLDR methods show five well-separated clusters (see
Figure~\ref{fig-gau1_sc}, Figure~\ref{fig-gau2_sc},
Figure~\ref{fig-gau3_sc}, Figure~\ref{fig-gau4_sc}, and
Figure~\ref{fig-gau5_sc}). This suggests that for the five Gaussian
cluster dataset, all NLDR methods effectively preserve the global
structure. tSNE displays clusters with varying densities, indicating
their ability to capture within-cluster variation (see
Figure~\ref{fig-gau1_sc}). On the other hand, both UMAP, PHATE, PaCMAP
and TriMAP show clusters with flat surfaces, suggesting a failure to
capture within-cluster variation (see Figure~\ref{fig-gau2_sc},
Figure~\ref{fig-gau3_sc}, Figure~\ref{fig-gau4_sc} and
Figure~\ref{fig-gau5_sc}). Therefore, UMAP, PHATE, PaCMAP and TriMAP do
not capture the local structure as effectively as other methods.

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_tsne_1.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_tsne_2.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_tsne_3.png}\end{minipage}%

\caption{\label{fig-gau1_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in \pD{},
a video of the tour animation is available at
(\url{https://youtu.be/RASEE7N5MbM}).}

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_umap_1.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_umap_2.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_umap_3.png}\end{minipage}%

\caption{\label{fig-gau2_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in \pD{},
a video of the tour animation is available at
(\url{https://youtu.be/iG4bCPkJilw}).}

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_phate_1.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_phate_2.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_phate_3.png}\end{minipage}%

\caption{\label{fig-gau3_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in \pD{},
a video of the tour animation is available at
(\url{https://youtu.be/L_PVLGwfOS0}).}

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_pacmap_1.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_pacmap_2.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_pacmap_3.png}\end{minipage}%

\caption{\label{fig-gau4_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
\pD{}, a video of the tour animation is available at
(\url{https://youtu.be/z07cKXi8EJQ}).}

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_trimap_1.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_trimap_2.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/five_gau_clusters/sc_trimap_3.png}\end{minipage}%

\caption{\label{fig-gau5_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
\pD{}, a video of the tour animation is available at
(\url{https://youtu.be/Chs1lYAoX2w}).}

When compare the NLDR representations and generated models, tSNE with
perplexity \(61\) appears to be a reasonable representation for
visualizing the five Gaussian cluster dataset. This is supported by
investigating the model generated with tSNE in the data space, which
provides evidence that it preserves both local and global structures.
Also, the NLDR representation with tSNE shows five well-separated
clusters.

@@

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_1.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_2.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_3.png}\end{minipage}%

@@

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_4.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_5.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_6.png}\end{minipage}%

@@

According to \textbf{?@fig-mnist1\_sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP (see
Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \pD{} data. Additionally, the model shows
a twisted pattern within the non-linear structure in \pD{} space (see
\textbf{?@fig-mnist1\_sc2}), which is an additional pattern not visible
in the \gD{} representation (see Figure~\ref{fig-pacmap-author}).
Furthermore, as shown in \textbf{?@fig-mnist1\_sc3}, some long edges
exist in the \pD{} space that are not recognized as long edges in the
\gD{} representation. However, PaCMAP is a \emph{good} \gD{}
representation of MNIST digit 1 data.

@@

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/mnist/sc_1.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/mnist/sc_2.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/mnist/sc_3.png}\end{minipage}%
